Multiple Dice Working as One: CAD Flows and Routing Architectures for Silicon Interposer FPGAs

### 2014-12-15 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
Fixed a typo.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4533 8e3573b8-cf2c-4f14-ef6d-137439e28b8b
updating all my graphs to be blakc&white friendly.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4535 8e3573b8-cf2c-4f14-ef6d-137439e28b8b
Updated my figures to look more like Ehsan's.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4537 8e3573b8-cf2c-4f14-ef6d-137439e28b8b
Updated square vs tall interposer figures to have the same area, compressed cad flow flowchart.
git-svn-id: https://vtr-verilog-t

% As a general rule, do not put math, special symbols or citations in the abstract or keywords.
\begin{abstract}
Interposer-based multi-FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. Such devices allow larger FPGA systems to be built than one monolithic die can accomodate and are now commercially available. An open question, however, is how efficient such systems are compared to a monolithic FPGA, as the number of signals passing between dice is reduced and the signal delay between dice is increased in an interposer system vs. a monolithic FPGA.

We create a new version of VPR to investigate the architecture of such systems, and show that by modifying the placement cost function to minimize the number of signals that must cross between dice we can reduce routing demand by 18\% and delay by 2\%. We also show that the signal count between dice and the signal delay between dice are key architecture parameters for interposer-based FPGA systems. We find that if an interposer supplies (between dice) 60\% of the routing capacity that the normal (within-die) FPGA routing channels supply, there is little impact on the routability of circuits. Smaller routing capacities in the interposer do impact routability however: minimum channel width increases by 20\% and 50\% when an interposer supplies only 40\% and 30\% of the within-die routing, respectively. The interposer also impacts delay, increasing circuit delay by 34\% on average for a 1 ns interposer signal delay and a four-die system. Reducing the interposer delay has a greater benefit in improving circuit speed than does reducing the number of dice in the system.
\end{abstract}

@@

\section{Introduction}
\label{introSection} 
%   - Describe multi-chip
Interposer-based multi-FPGA systems are composed of multiple FPGA dice, which are connected through a silicon interposer. The interposer is fabricated with an older technology than that used for the dies, and links between dice include both a micro-bump on each FPGA die and a metal wire on the interposer~\cite{xilinxTSVperformance}. This results in a reduced connectivity between dice and increased delay for connections between dice, as compared to the total routing connectivity and the delay one can achieve within a single die. In this paper we present an architecture study of silicon interposer-based FPGAs that analyzes how the reduced connectivity and increased delay impact their performance.

%   - Motivate: yield, bigger FPGA, no previous study
The silicon interposer FPGA technology is interesting because it enables the creation of large FPGAs composed of small dies and also of very large FPGAs, with higher logic capacity than one can achieve with a single die. Such FPGA systems are sometimes called 2.5D FPGAs, since they make use of vertical stacking of dies on an interposer to enable higher integration levels.

% yield rate
Being able to make large FPGAs with multiple smaller dies is particularly interesting at the beginning of a new manufacturing process, when defect densities are high. In such a case, good-die yield drops dramatically as the die size increases, and this drastically impacts the availability of large FPGAs early in the process lifetime.

%   - Prior work (El Gamal, Xilinx WP, Xilinx paper) - TODO
The idea of using 3 dimensions to design FPGAs is not new. \cite{3dfpga1995} proposed the creation of 3D FPGAs by stacking 2D FPGAs and connecting them with solder bumps. Lin et al also studied 3D FPGAs, but using a different approach~\cite{3dfpga}, proposing multiple active layers, with different FPGA functions in each. While promising, both approaches have manufacturing challenges: \cite{3dfpga1995} requires a higher density of through-silicon vias than can currently be manufactured and the multiple active layers required by \cite{3dfpga} are not yet widely available. In contrast, silicon interposers linked to dies via microbumps are now manufacturable.

@@

%   - Contributions
The main contributions of our work are as follows:
\begin{itemize}
%     * Study on routability vs wires cut
	\item A detailed study of how the multi-FPGA system routability is impacted by the amount of connectivity between dice provided by the interposer.
%     * Timing vs wires cut + delay
	\item An Analysis of the impact of the interposer on timing.
%     * CAD flow optimizations for 2.5D FPGAs
	\item Modification of the VPR CAD tool to model and target 2.5D silicon interposer-based FPGAs, as well as CAD changes to optimize for 2.5D FPGAs.
%     * Analysis of existing Virtex-7 FPGAs
	\item Analysis of the commercially available Xilinx Virtex-7 silicon interposer FPGAs~\cite{xilinxWP}\cite{xilinx7series}. We measure the connectivity reduction and delay increase between dies caused by the interposer on such FPGAs in order to see where these devices lie within the architcture space we explore.
\end{itemize}

The paper is organized as follows. Section \ref{siliconSection} gives more information about silicon interposer technology. Section \ref{archSection} describes the changes made to the FPGA architecture description and to VPR to target 2.5D FPGAs. Section \ref{cadSection} describes optimizations made to the VPR placement algorithm to improve quality on such FPGAs and Section \ref{resultsSection} presents architectural results for 2.5D FPGAs.

@@

% 1 - FPGAs "More than Moore" - want(?) to combine chips more tightly / with more connectivity than on a board
As mentioned in the previous section, interposer-based FPGAs allow the creation of chips larger than a single die, making a "More than Moore" improvement on the size and number of logic elements possible, and with chips combined far more tightly and with more connectivity than if they were on separate boards connected through conventional means.

% 2 - a) - Can sell bigger chips
%        - Can we make bigger? -> 28nm 1.95m LE vs 979k (V7) / 952k (SV)
The improvement in the number of logic elements of 2.5D FPGAs over conventional ones is very significant. Xilinx's largest interposer-based FPGA, the Virtex-7 XC7V2000T, has 4 dies (which Xilinx calls Super Logic Regions) and 1.954 million logic elements~\cite{xilinx7series}, while the largest non-interposer Virtex-7 die (the XC7VX980T), has 979k logic elements and Altera's largest FPGA, the Stratix V 5SGXBB, has 952k logic elements~\cite{stratixV}. Even though all these FPGAs use a 28 nm process, silicon interposer technology allows the creation of FPGAs with twice the resources possible on even an extremely large single die.

%     b) - Yield results
%               -> Ref paper / use lecture slides
%               -> Defect density of 1/cm^2 is typical at early days of new process
%               -> FPGA early adopters of new process
Another major advantage of interposer-based FPGAs comes at the beginning of a new manufacturing process, when the defect density is high~\cite{xilinxTSVperformance}. Bigger dice suffer a much reduced yield compared to smaller dice, and this greatly affects the supply and cost of top-of-the-line FPGAs to early adopters.

To illustrate this impact, consider a new process in which the defect density is 1/$cm^2$, which is a reasonable value early in the process lifecycle, and the die area is $6cm^2$, which roughly matches the size of the largest member of a high-end FPGA family such as Virtex 7. Using the Poisson Yield Model~\cite{yieldmodel}, the yield is only $0.25\%$ of die. If instead the chip is composed of four $1.5cm^2$ dies, the yield is $22\%$. This means that a 12 inch silicon wafer with $730cm^2$ of area would produce on average $0.3$ working $6cm^2$ dies, while the same wafer would produce on average $107$ working $1.5cm^2$ dies. Therefore, as a $6cm^2$ chip would be composed of four $1.5cm^2$ dies, the wafer would yield $26.75$ systems on average, as the "assembly yield" of placing these four die on an interposer is very high~\cite{xilinxTSV}. Hence the number of interposer-based FPGAs created from the same silicon wafer would be almost $100\times$ greater than a monolithic FPGA of the same size.

@@

The Xilinx 2.5D FPGAs from the Virtex-7 family are currently the only commercially available silicon interposer-based FPGAs. As described in Section~\ref{archSection} we are studying the impact of several key interposer parameters on the performance of the multi-FPGA system, including (i) the percentage of the wiring normally present between rows of the FPGA which are cut when crossing between dice, and (ii) the extra delay (vs. a normal connection between adjacent rows) added when one must traverse the interposer.  To locate where Virtex-7 lies in this architecture space, we combined published information on the implementation of Xilinx's interposer-based FPGAs~\cite{xilinxTSV} with a detailed analysis of the~\textit{XC7V2000T} FPGA routing resources visible in the \textit{Vivado Device View}.

@@

The 28nm dies are connected to the 65nm silicon interposer through microbumps with a 45$\mu$m pitch. Hence the area occupied by microbumps at one edge of one die is 13440 $\times$ (45$\mu$m)\textsuperscript{2} $=$ 27mm\textsuperscript{2}. If we assume each die is $7 \times 12 mm$, as presented by Chaware et al. in \cite{xilinxTSV}, the bumps have to be spread out near the edge and need to go as far as $2.25mm$ away from the edge of the die. This greater distance from the border increaes the length of the inter-die connections, and along with the presence of the micro-bumps and their capacitance, leads to an increased delay for these crossing wires vs. that of a typical on-die routing wire. Chaware et al. state that the latency to cross the interposer is approximately $1ns$. For comparison, a typical medium length 28 nm FPGA routing wire (e.g. spanning four logic blocks) has a delay of approximately 125 ps, while a longer wire (e.g. spanning 12 logic blocks) has a delay of approximately 250 ps.

Overall, these interposer-based FPGAs have increased delay and reduced connectivity between dies, with approximately $23\%$ of the usual number of vertical wires crossing between dies and approximately $1ns$ of increased delay to cross the interposer.

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture Models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{fpga.eps}
\caption{A sample two-die / 1-cutline architecture containing both logic blocks and RAM blocks.}
\label{fig:fpga}
\end{figure}

@@

%   - Interposer-Routing interface
\subsection{Interposer-Routing Interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

@@

As illustrated in Figure~\ref{fig:vpr_interposer_gui}, no vertical routing wires cross a cutline boundary and a fraction of them connect to other dies via interposer wires.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{bb_new.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{arch_experiments3.eps}
\caption{Impact of routing-interposer interface flexibility on minimum channel width and circuit speed.}
\label{fig:all_arch_experiments}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{wires_cut_4part.eps}
\caption{Impact of number of dice on routability. TODO JAVEED -- LU32 90\% result is missing}
\label{fig:wires_cut_4part}
\end{figure}

Systems with more dice are slower, on average the critical path delay is xx\% higher for a four-die system than a two-die system. This speed difference is not significantly affected by the \textit{\% wires cut}.
 TODO JAVEED -- figure out critical path delay %.

@@

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[height = 5cm]{square_fpga.pdf}}
\subfigure[]{\includegraphics[height = 5cm]{tall_fpga.pdf}}
\caption{Devices with (a) a square interposer and (b) square dice.}
\label{fig:square_tall}
\end{figure}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Conclusions and Future Work}
\label{conclusionSection}
We have extended VPR to model and optimize for interposer-based multi-FPGA systems. While such systems are now a commercial reality, we do not know of any prior study of their key architectural parameters. We found that by modifying VPR's placement cost function we could improve the routability (reduce minW) by 18\%, while simultaneously improving speed by 2\%. Interestingly, we found that a smooth cost function that less precisely models the interposing wiring demand outperformed direct monitoring of the number of required interposer connections during annealing.

We defined three key architecture parameters for interposer-based FPGAs, and used this extended VPR to analyze their impact on minimum channel width and critical path delay. We found that the \textit{minimum channel width} increases steeply after more than 60\% of the within-die wires are cut at the interposer, and when 80\% of wires are cut the \textit{minimum channel width} is more than double that required by a monolithic FPGA. The Xilinx \textit{XC7V2000T} FPGA has 77\% of the normal wiring cut at the interposer, indicating that commercial interposer-based FPGAs will require high-quality CAD optimization to maintain good routability. The \textit{critical path delay} is not strongly influenced by the \textit{\% wires cut} but is strongly influenced by the interposer delay and the number of cuts. Our results show that the critical path usually crosses the interposer cutline more than once, as the \textit{critical path delay} is increased by more than the interposer \textit{delay increase}. Increasing the number of dice in an interposer-based FPGA does not significantly impact the \textit{minimum channel width}, but does lead to a larger \textit{critical path delay}.

There are many interesting CAD and architecture questions for interposer-based FPGAs which we plan to investigate.  Currently the wires that cross the interposer are accessed with the same switch structure as other wires in the FPGA. Since interposer wires are more scarce, possibly these wires should have larger multiplexers feeding them, which would make them easier to use; this may help routability of the system. Such changes will have to be carefully considered however, to make sure the FPGA can still be laid out as an array of regular tiles. Alternative CAD flows to improve the interposer routability are also possible. For example, instead of following the synthesize, place and route CAD flow of a conventional FPGA, we could add a partitioning step before placement which would divide the circuit into one partition per die. Such a flow may improve routability, as a partitioner's main focus is minimizing the number of cut signals. Enhanced partitioners will be required, as current partitioners such as hMetis~\cite{hMetis} cannot model the heterogeneous balance constraint (i.e. use no more than the available device logic, RAM, DSP and I/O blocks in any die) present in FPGAs. Additionally, by dividing the placement problem into two pieces -- partitioning and within-die placement -- we may increase the critical path delay as we can no longer globally optimize the placement of timing critical paths in one unified placement step. Nonetheless, this is a viable alternative approach and a comparison to the approach taken in this paper would be very interesting.

>>>>

% As a general rule, do not put math, special symbols or citations in the abstract or keywords.
\begin{abstract}
Multiple FPGA dice can be connected through a silicon interposer to form a larger FPGA system than one monolithic die can accomodate. Such systems are now commercially available but many open questions remain concerning their architecture and efficiency, as the number of signals passing between dice is reduced and the signal delay between dice is increased compared to a monolithic FPGA.
We create a new version of VPR to target such systems and investigate modifications to placement and routing and the incorporation of partitioning into the flow to best exploit interposer-based FPGAs. Our best CAD flow can reduce the routing demand for interposer FPGAs with realistic amounts of connectivity between dice by 47\% and improve the circuit speed by 13\% on average. Modifications to the FPGA architecture to add routing flexibility when crossing the interposer are very beneficial and can improve routability by a further 11\%. With these CAD and architecture enhancements, we find that if an interposer supplies (between dice) 40\% of the routing capacity that the normal (within-die) FPGA routing channels supply there is negligible impact on the routability of circuits. Smaller interposer routing capacities do impact routability however: minimum channel width increases by 10\% and 70\% when an interposer supplies only 20\% and 10\% of the within-die routing, respectively. The interposer also impacts delay, increasing circuit delay by 11\% on average for a 1 ns interposer signal delay and a two-die system. Reducing the interposer delay has a greater benefit in improving circuit speed than does reducing the number of dice in the system.
\end{abstract}

@@

\section{Introduction}
\label{introSection} 
%   - Describe multi-chip
Interposer-based FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. The interposer is fabricated with an older technology than that used for the dice, and links between dice include both a micro-bump on each FPGA die and a metal wire on the interposer~\cite{xilinxTSVperformance}. This results in a reduced connectivity between dice and increased delay for connections between dice compared to the total routing connectivity and the delay one can achieve within a single die. In this work we present an architecture study of silicon interposer-based FPGAs that analyzes how the reduced connectivity and increased delay impact their performance.

%   - Motivate: yield, bigger FPGA, no previous study
Silicon interposer FPGA technology is interesting because it enables the creation of large FPGAs composed of small dice and also of very large FPGAs, with higher logic capacity than one can achieve with a single die. Such interposer systems are sometimes called 2.5D devices, since they make use of vertical stacking of dice on an interposer to enable higher integration levels~\cite{3dhandbook}.

% yield rate
Making large FPGAs from multiple smaller dice is particularly interesting at the beginning of a new manufacturing process, when defect densities are high. In such a case, good-die yield drops dramatically as the die size increases, and this drastically impacts the availability of large FPGAs early in the process lifetime.

%   - Prior work (El Gamal, Xilinx WP, Xilinx paper) - TODO
The idea of using 3 dimensions to design FPGAs is not new. \cite{3dfpga1995} proposed the creation of 3D FPGAs by stacking 2D FPGAs and connecting them with solder bumps. Lin et al also studied 3D FPGAs, but using a different approach~\cite{3dfpga} that utilized multiple active layers with different FPGA functions (logic, routing, and configuration) in each. While promising, both these approaches have manufacturing challenges: \cite{3dfpga1995} requires a higher density of through-silicon vias than can currently be manufactured and the multiple active layers required by \cite{3dfpga} are not yet widely available. In contrast, silicon interposers linked to dice via microbumps are now manufacturable.

@@

%   - Contributions
The main contributions of our work are as follows:
\begin{itemize}
	\item Modifications of the VPR CAD tool to model and target 2.5D silicon interposer-based FPGAs, as well as CAD changes to optimize for 2.5D FPGAs.
	\item An investigation of the efficacy of incorporating partitioning into various stages of the CAD flow.
	\item A detailed study of how the multi-FPGA system routability is impacted by the amount of between-die connectivity provided by the interposer, and by the switching flexibility at the FPGA die - interposer boundary.
	\item An analysis of the interposer's impact on timing.
\end{itemize}

The paper is organized as follows. Section \ref{siliconSection} gives more information about silicon interposer technology. Section \ref{archSection} describes the changes made to the FPGA architecture description and to VPR to target 2.5D FPGAs. Section \ref{cadSection} describes optimizations made to the VPR placement and routing algorithms to improve quality on such FPGAs and Section~\ref{sec:CADflow} describes how partitioning can be used to further enhance the CAD flow. Section \ref{resultsSection} presents architectural results for 2.5D FPGAs.

An earlier and less detailed version of this work appeared in \cite{interposer2014}. We extend this earier work by enhancing the CAD tools, incorporating partitioning into the flows and investigating a broader range of architecture questions.

@@

Interposer-based FPGAs allow the creation of chips larger than a single die, making a "More than Moore" improvement on the size and number of logic cells possible, and with chips combined far more tightly and with more connectivity than if they were on separate boards connected through conventional means.

The improvement in the capacity of 2.5D FPGAs over conventional ones is very significant. Xilinx's largest 28 nm interposer-based FPGA, the Virtex-7 XC7V2000T, contains 4 FPGA dice (which Xilinx calls Super Logic Regions) and 1.954 million logic cells~\cite{xilinx7series}, while the largest non-interposer Virtex-7 die (the XC7VX980T), has 979k logic cells and Altera's largest 28 nm FPGA, the Stratix V 5SGXBB, has 952k logic elements~\cite{stratixV}. Even though all these FPGAs use a 28 nm process, silicon interposer technology allows the creation of FPGAs with twice the resources possible on even an extremely large single die. Xilinx's recently announced Virtex UltraScale line of 20 nm FPGAs makes even more aggressive use of silicon interposers; the largest monolithic FPGA in this family contains 941k logic cells, while the four largest capacity FPGAs are all built with silicon interposers and span a range from 1.25 to 4.43 million logic cells~\cite{virtexultrascale}.

Another major advantage of interposer-based FPGAs comes at the beginning of a new manufacturing process, when the defect density is high~\cite{xilinxTSVperformance}. Bigger dice suffer a much reduced yield compared to smaller dice, and this greatly affects the supply and cost of top-of-the-line FPGAs to early adopters.

To illustrate this impact consider a new process in which the defect density is 1/$cm^2$, which is a reasonable value early in the process lifecycle, and the die area is $6cm^2$, which roughly matches the size of the largest member of a high-end FPGA family such as Virtex 7. Using the Poisson Yield Model~\cite{yieldmodel}, the yield is only $0.25\%$ of die. If instead the chip is composed of four $1.5cm^2$ dice, the yield is $22\%$. This means that a 12 inch silicon wafer with $730cm^2$ of area would produce on average $0.3$ working $6cm^2$ dice, while the same wafer would produce on average $107$ working $1.5cm^2$ dice. Therefore, as a $6cm^2$ chip would be composed of four $1.5cm^2$ dice, the wafer would yield $26.75$ systems on average, as the "assembly yield" of placing these four die on an interposer is very high~\cite{xilinxTSV}. Hence the number of interposer-based FPGAs created from the same silicon wafer would be almost $100\times$ greater than a monolithic FPGA of the same size.

@@

The Xilinx 2.5D FPGAs from the Virtex-7 and Virtex Ultrascale families are currently the only commercially available silicon interposer-based FPGAs~\cite{xilinxWP}. As described in Section~\ref{archSection} we are studying the impact of several key interposer parameters on the performance of the multi-FPGA system, including (i) the percentage of the wiring normally present between rows of the FPGA which are cut when crossing between dice, and (ii) the extra delay (vs. a normal connection between adjacent rows) added when one must traverse the interposer.  To locate where Virtex-7 lies in this architecture space, we combined published information on the implementation of Xilinx's interposer-based FPGAs~\cite{xilinxTSV} with a detailed analysis of the~\textit{XC7V2000T} FPGA routing resources visible in the \textit{Vivado Device View}.

@@

The 28nm dice are connected to the 65nm silicon interposer through microbumps with a 45$\mu$m pitch. Hence the area occupied by microbumps at one edge of one die is 13440 $\times$ (45$\mu$m)\textsuperscript{2} $=$ 27mm\textsuperscript{2}. If we assume each die is $7 \times 12 mm$, as presented by Chaware et al. in \cite{xilinxTSV}, the bumps have to be spread out near the edge and need to go as far as $2.25mm$ away from the edge of the die. This greater distance from the border increaes the length of the inter-die connections, and along with the presence of the micro-bumps and their capacitance, leads to an increased delay for these crossing wires vs. that of a typical on-die routing wire. Chaware et al. state that the latency to cross the interposer is approximately $1ns$. For comparison, a typical medium length 28 nm FPGA routing wire (e.g. spanning four logic blocks) has a delay of approximately 125 ps, while a longer wire (e.g. spanning 12 logic blocks) has a delay of approximately 250 ps.

Overall, these interposer-based FPGAs have increased delay and reduced connectivity between dice, with approximately $23\%$ of the usual number of vertical wires crossing between dice and approximately $1ns$ of increased delay to cross the interposer.

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture Models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are used as-is, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.65\linewidth]{fpga.eps}
\caption{A sample two-die / 1-cutline architecture containing both logic blocks and RAM blocks.}
\label{fig:fpga}
\end{figure}

@@

%   - Interposer-Routing interface
\subsection{Interposer-Routing Interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires at the edge of an FPGA die that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

@@

As illustrated in Figure~\ref{fig:vpr_interposer_gui}, no vertical routing wires cross a cutline boundary and a fraction of them connect to other dice via interposer wires.

@@

\begin{figure}[!h]
\centering
\includegraphics[width=0.75\linewidth]{bb_new.eps}
\caption{Illustration of three different scenarios for the wiring cost. The dashed green box shows a case where \textit{(a)} it crosses the interposer, the dotted black box shows a case where \textit{(b)} it barely crosses the cutline, and the solid blue one shows a case where \textit{(c)} the bounding box does not span the cutline.}
\label{fig:bb_illustration}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{arch_experiments4.eps}
\caption{Impact of routing-interposer interface flexibility on minimum channel width and circuit speed.}
\label{fig:all_arch_experiments}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\input{wires_cut_4part}
\caption{Impact of number of dice on routability. TODO JAVEED -- LU32 90\% result is missing}
\label{fig:wires_cut_4part}
\end{figure}

Systems with more dice are slower; on average the critical path delay is xx\% higher for a four-die system than a two-die system. This speed difference is not significantly affected by the \textit{\% wires cut}.
 TODO JAVEED -- figure out critical path delay %.

@@

\begin{figure}[!htbp]
\centering
\subfigure[]{\includegraphics[height = 5cm]{square_fpga.pdf}}
\subfigure[]{\includegraphics[height = 9.335cm]{tall_fpga.pdf}}
\caption{Devices with (a) a square interposer and (b) square dice.}
\label{fig:square_tall}
\end{figure}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Conclusion}
\label{conclusionSection}
We have extended VPR to model and optimize for interposer-based multi-FPGA systems. While such systems are now a commercial reality, we do not know of any prior study of their key architectural parameters. We found that by modifying VPR's placement cost function we can improve routability (reduce minW), while simultaneously improving speed. Modifications to the VPR router lookahead function are crucial to efficiently routing interposer-based FPGAs, yielding a 31$\times$ speed-up. Incorporating partitioning into the CAD flow is very beneficial. Our best partitioning flow converts nets (hyperedges) to a star graph model, uses Metis to partition the resulting circuit graph, and forces VPR's packer to respect the Metis partitioning to FPGA dice. Taken together, these CAD changes reduce the signal wires that must cross the interposer by an average of 47\% while simultaneously improving circuit speed by 13\% for realistic interposer architectures.

We defined four key architecture parameters for interposer-based FPGAs, and used this extended VPR to analyze their impact on minimum channel width and critical path delay. We find that adding a moderate amount of switching flexibility at the FPGA - interposer boundary has negligible area cost but reduces the required FPGA minimum channel width by 13\%. We also find that the multi-die system has good routability so long as the interposer provides sufficient routing bandwidth, and this routing bandwidth can be much less than one would find in a monolithic FPGA. Specifically, if the interposer provides 40\% or more of the channel width required by the monolithic FPGA (an average of 45 wires per channel across our benchmark suite) the routability of the interposer-based FPGA is equivalent to that of a monolithic one. As the interposer signal bandwidth becomes more constrained, there is some impact to routability, but even an interposer providing only 20\% of the monolithic FPGA routing channel bandwidth (22 wires in our experiments) only needs a 10\% increase in the within-FPGA-die routing channel width to compensate.

The \textit{critical path delay} is not strongly influenced by the \textit{\% wires cut} but is strongly influenced by the interposer delay and the number of cuts. Our results show that the critical path often crosses the interposer cutline more than once, as the \textit{critical path delay} is increased by more than the interposer \textit{delay increase}; on average we find that a 1 ns interposer delay causes an 11\% delay increase in a two-die system vs. a monolithic FPGA . Increasing the number of dice in an interposer-based FPGA does not greatly impact the \textit{minimum channel width}, but does lead to a larger \textit{critical path delay}. We also found that the aspect ratio of the interposer-based FPGA is an important architectural parameter; for the best routability one should use a square interposer and rectangular FPGA dice.

Overall we find the results for interposer-based FPGAs to be very positive. While an interposer can only provide a fraction of the signal count between dice that one would have across cutline within a die, the system still has excellent routability and behaves similarly to a monolithic FPGA so long as the interposer signal count is not pushed to very low values; approximately 15\% of the within-die routing in our experiments. We have also found that modifications to the FPGA CAD flow and routing architecture to exploit interposer-based FPGAs have been very productive. As this is a new area we believe more routability and timing enhancements are likely with further research, and this would further enhance the utility of interposer-based FPGAs.
